---
title: "Test data generation"
output:
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: lumen
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create education distribution

Generate probabilty distribution for education

```{r warning=FALSE,message=FALSE}

#library with random data generator
library(wakefield)
library(generator)
library(sqldf) 
library(readr)
library("rjson")

#https://www.census.gov/data/tables/2019/demo/educational-attainment/cps-detailed-tables.html
#All Races table 2019 distribution
#https://www.tutorialspoint.com/r/r_data_frames.htm
#https://nces.ed.gov/programs/coe/pdf/coe_cba.pdf

 
 
# Create the data frame of code table and probability  ratio to generate ramdom data.
educationCodeDf <- data.frame(
  education = c(
    "Not reached high school",
    "High school dropput",
    "High school graduate",
    "Some college, no degree",
    "Associate's degree, occupational",
    "Associate's degree, academic",
    "Bachelor's degree",
    "Master's degree",
    "Professional degree", #(lawyer,dr)
    "Doctoral degree"
  ),
  studentCnt = c(8879, 17680, 70947, 45028, 10381, 14168, 53312, 22459, 3150, 4557),
  code         =c("NRHS",  "HSDP",  "HSGR",  "SCND",  "ASDO",   "ASDA",  "BSDG",  "MSDG",  "PRDG",   "DRDG"  ),
  SalaryLower  = c(20000,   30000,   35000,   40000,   60000,   60000,   70000,   90000,   150000,   130000  ),
  SalaryHigher = c(30000,   50000,   65000,   70000,   90000,   90000,   95000,   120000,  250000,   180000  )
)
educationCodeDf$studentratio=educationCodeDf$studentCnt/sum(educationCodeDf$studentCnt)
#Generate random list of education from the above records with probabilty studentratio
education  <- education(100, x = educationCodeDf$code, prob = educationCodeDf$studentratio, name = "Education")
print(education)

#Convert the list into dataframe and append to personDf
personDf <- as.data.frame(education)
```
##  Create state code data frame

State code dataframe for state address 

```{r warning=FALSE,message=FALSE}

StateCodeDF = data.frame (StateCode =c("Alabama - AL",
                                       "Alaska - AK",
                                       "Arizona - AZ",
                                       "Arkansas - AR",
                                       "California - CA",
                                       "Colorado - CO",
                                       "Connecticut - CT",
                                       "Delaware - DE",
                                       "Florida - FL",
                                       "Georgia - GA",
                                       "Hawaii - HI",
                                       "Idaho - ID",
                                       "Illinois - IL",
                                       "Indiana - IN",
                                       "Iowa - IA",
                                       "Kansas - KS",
                                       "Kentucky - KY",
                                       "Louisiana - LA",
                                       "Maine - ME",
                                       "Maryland - MD",
                                       "Massachusetts - MA",
                                       "Michigan - MI",
                                       "Minnesota - MN",
                                       "Mississippi - MS",
                                       "Missouri - MO",
                                       "Montana - MT",
                                       "Nebraska - NE",
                                       "Nevada - NV",
                                       "New Hampshire - NH",
                                       "New Jersey - NJ",
                                       "New Mexico - NM",
                                       "New York - NY",
                                       "North Carolina - NC",
                                       "North Dakota - ND",
                                       "Ohio - OH",
                                       "Oklahoma - OK",
                                       "Oregon - OR",
                                       "Pennsylvania - PA",
                                       "Rhode Island - RI",
                                       "South Carolina - SC",
                                       "South Dakota - SD",
                                       "Tennessee - TN",
                                       "Texas - TX",
                                       "Utah - UT",
                                       "Vermont - VT",
                                       "Virginia - VA",
                                       "Washington - WA",
                                       "West Virginia - WV",
                                       "Wisconsin - WI",
                                       "Wyoming - WY" ))
 
```
</p>
##  Employment type reference table

Reference data for employment
```{r  warning=FALSE,message=FALSE}
# Create employment reference table with probability  distribution 
employmentCodeDf <- data.frame(
  employment = c(
    "Full time",    "Part time",    "Unemployed",    "Retired"   ),
  employmentRatio = c(0.55, 0.15, 0.1,  0.2),
  code = c(
    "FT", "PT", "UNEMP", "REIT"
  )
)

# Using employment reference table create random list of codes and combine code to personDf
emp <- employment(100, x = employmentCodeDf$code, 
                  prob = employmentCodeDf$employmentRatio, name = "Employment")
employment <- as.data.frame(emp)
personDf$emptstatus <- employment$emp
 
```
</p>
##  Birrthdate data

Populate birth dates and age
```{r}
# Create employment reference table with probability  distribution 
employmentCodeDf <- data.frame(
  employment = c(
    "Full time",    "Part time",    "Unemployed",    "Retired"   ),
  employmentRatio = c(0.55, 0.15, 0.1,  0.2),
  code = c(
    "FT", "PT", "UNEMP", "REIT"
  )
)

# Using employment reference table create random list of codes and combine code to personDf
emp <- employment(100, x = employmentCodeDf$code, 
                  prob = employmentCodeDf$employmentRatio, name = "Employment")
employment <- as.data.frame(emp)
personDf$emptstatus <- employment$emp


# Create random birthdates for all population in reiterment
## Get birthdate for retired
dobrt <- as.Date(r_date_of_births(100, start = Sys.Date() - 365 * 90, end = Sys.Date() - 365 * 65))
dobrtdf <- as.data.frame(dobrt)
## Get birthdate for non retired
dobnrt <- as.Date(r_date_of_births(100, start = Sys.Date() - 365 * 65, end = Sys.Date() - 365 * 20))
dobnrtdf <- as.data.frame(dobnrt)

#Get data type of dataframe
#sapply(dobrt, typeof)


# Add RowID to dataframes to combine based on rowID and other attributes
personDf$ID <- seq.int(nrow(personDf))
dobrtdf $ID <- seq.int(nrow(dobrtdf))
dobrtdf $emp = "REIT"
dobnrtdf$ID <- seq.int(nrow(dobnrtdf))


# Combine Birthdate and salary range based on all data frame and conditions of education
personDf = sqldf("
  SELECT d2.dobrt,d1.*  ,d3.dobnrt,SalaryLower,SalaryHigher
  FROM personDf d1 
  INNER JOIN educationCodeDf ecdf
    on d1.education =ecdf.code
  LEFT OUTER JOIN dobrtdf d2
  ON d1.ID = d2.ID
  and d1.emptstatus    =d2.emp
  LEFT OUTER JOIN dobnrtdf d3
  ON d1.ID = d3.ID
  and  emptstatus != 'REIT'
  
")

#personDf had 2 dob columns coming from 2 dataframe based on emp status , 
#combine those columns into one and delete other 2 columns
#Also calculate person age
personDf$DOB <-  ifelse(is.na(personDf$dobrt), personDf$dobnrt, personDf$dobrt) 
personDf$DOB <- as.Date(personDf$DOB, origin="1970-01-01")  
personDf$age <- as.integer(difftime(Sys.Date(), personDf$DOB, units='days'))
personDf$age <- personDf$age/365
```
</p>
##  Gender , StateCode, ZIp

Statecode and Zipcode are not aligned as per US postal services. Needs more work
```{r  warning=FALSE,message=FALSE}

# Assign gender to these records 
Gender <- sex(100, x = c("Male", "Female"), prob = c(0.52,
                                                     0.48), name = "Gender")
personDf$Gender <- as.vector(unlist(Gender))

#Assign zip code
zip_code <- zip_code(100, k = 10, x = 10000:99999, prob = NULL, name = "Zip")
personDf$ZipCode <- as.vector(unlist(zip_code))
State_Code <- employment(100, x = StateCodeDF$StateCode, prob = NULL,  name = "StateCode")
personDf$State <- as.vector(unlist(State_Code))

```
</p>
##  Income and other info

For each record create income based on range and assign first name, last name, drop unused columns
```{r  warning=FALSE,message=FALSE}

#For each record between lower bound and higher bound calculate  random salary. runif has to be executed at each record level
for (row in 1:nrow(personDf)) {
  personDf[row, "Sal"]=  runif(1,  personDf[row, "SalaryLower"],    personDf[row, "SalaryHigher"]) 
  
}

#Assign other information full name,email,SSN,CardNumber
personDf$FullName <- generator::r_full_names(100)
personDf$email <-  generator::r_email_addresses(100)
personDf$TIN <- generator::r_national_identification_numbers(100)
personDf$CardNumber <- generator::r_credit_card_numbers(100)

personDf$dobrt      <- NULL
personDf$dobnrt     <- NULL


#### Address info can be modified using following data source

# Give the input file name to the function. Address improvements 
#http://results.openaddresses.io/
addressjson <- fromJSON(file = 'https://raw.githubusercontent.com/EthanRBrown/rrad/master/addresses-us-100.json')
# Convert JSON file to a data frame.
addressDF <- as.data.frame(addressjson)
addressDF$ID <- seq.int(nrow(addressDF))


print(personDf)
